---
title: "Extended analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{extended_analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.height = 4,
  dpi = 300
)
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(outbreaker2)
library(o2ools)
```

# Reproduction number

We can estimate the case reproductive number ($R_i$), *i.e.* the number of secondary infections generated by each case, with the `get_Ri` function:

```{r}
out_id <- identify(out, linelist$name)

# A matrix of R values
# Columns refer to the case ID
# Rows refer to the MCMC iteration
Ri_mat <- get_Ri(out_id)
```

You can plot the offspring distribution:

```{r}
Ri_mat %>% 
  mutate(step = row_number()) %>% 
  pivot_longer(-step, names_to = "name", values_to = "Ri") %>%
  group_by(step, Ri) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup() %>%
  group_by(Ri) %>%
  summarise(
    mean = mean(percent),
    lower = quantile(percent, 0.025),
    upper = quantile(percent, 0.975),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = Ri, y = mean)) +
  geom_segment(aes(x = Ri, xend = Ri, y = 0, yend = mean), colour = "grey60", size = 0.5) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(x = "Number of secondary infections (Ri)",
       y = "Mean % of cases",
       title = "Posterior offspring distribution with 95% credible intervals")+
  theme_bw()
```

# Serial interval

The serial interval (SI) is the time between the onset of symptoms in an infector-infectee pair. You can compute the SI's empirical cumulative density functionusing with `get_si()` :

```{r}
trees <- get_trees(out_id, onset = linelist$onset)
si <- get_si(trees, date_suffix = "onset")

si |>
  ggplot(aes(x = x, y = mean))+
  geom_point() +
  geom_ribbon(aes(ymin = lwr, ymax = upr),
              fill = "grey",
              alpha = 0.5) +
  labs(x = "serial interval", y = "Percent")+
  theme_bw()
```

# Model convergence

*outbreaker2* uses Bayesian inference to account for uncertainty in who infected whom by sampling multiple plausible transmission trees from the posterior distribution. Standard MCMC diagnostics evaluate parameter chains (see `?plot.outbreaker_chains` and `?coda::gelman.diag`) rather than transmission events themselves.

[*mixtree*](https://cygei.github.io/mixtree/) provides a statistical framework for comparing collections of transmission trees (*epidemic forests*) to assess whether they originate from the same generative process. To evaluate MCMC convergence, *mixtree* can be used to compare epidemic forests sampled from multiple parallel MCMC chains. If the chains have converged, the resulting epidemic forests should be statistically indistinguishable.

## Run multiple chains of *outbreaker2*

```{r}
data <- outbreaker_data(dna = fake_outbreak$dna, dates = linelist$onset, ctd = fake_outbreak$ctd, w_dens = fake_outbreak$w)

# Run multiple chains of outbreaker2 in parallel with furrr
library(furrr)
n_chains <- 5
plan(multisession, workers = n_chains) #make sure to set the number of workers to the number of cores on your machine

set.seed(123)
chains <- future_map(1:n_chains, function(i) {
  out <- outbreaker(data = data)
  #brunin
  out[out$step>500, ]
}, .options = furrr_options(seed = TRUE))

```

## Extract trees

```{r}
trees <- lapply(chains, function(x) {
  get_trees(x)
})

```

We now have 5 epidemic forests, one for each chain.

#### Chi-squared test

```{r}
library(mixtree)

set.seed(123)
do.call(what = mixtree::tree_test, args = c(trees, list(
  method = "chisq",
  test_args = list(simulate.p.value = TRUE, B = 999)
)))
```

We used the chi-squared test to test the null hypothesis that the frequency of infector-infectee pairs is the similar between forests. A p-value of 1 indicates similarity between the forests, suggesting convergence. The chi-squared test allows for multiple introductions, unlike PERMANOVA.

#### PERMANOVA

PERMANOVA only accepts one introduction event, so we need to re-run *outbreaker2* allowing for a single introduction event only. We can do this by setting the `find_import` argument to `FALSE`.

```{r}
set.seed(123)
chains <- future_map(1:n_chains, function(i) {
  out <- outbreaker(
    data = data,
    config = outbreaker2::create_config(find_import = FALSE, move_kappa = FALSE)
  )
  #brunin
  out[out$step>500, ]
}, .options = furrr_options(seed = TRUE))

trees <- lapply(chains, function(x) {
  tree_list <- get_trees(x)
  lapply(tree_list, function(tree) {
    # Remove the imported case
    tree[!is.na(tree$from), ]
  })
})

set.seed(123)
do.call(what = mixtree::tree_test, args = c(trees, list(
  method = "permanova",
  test_args = list(permutations = 100)
)))
```

PERMANOVA tests whether the variance in tree topology is similar between *forests*. A p-value \> 0.05 suggests that the topologies of the trees do not differ significantly across the 5 *forests*, suggesting convergence.

# Group transmission patterns

Transmission chains can inform on the patterns of transmission between groups. [*linktree*](https://cygei.github.io/linktree/) provides a framework for estimating *group transmission assortativity* which quantifies the extent to which individuals transmit within their own group compared to others.

This requires knowledge of group sizes or their relative proportions. In our example, say the staff-to-patient ratio in the hospital is 1:3. It is also advised to analyse the transmission chains before *saturation* (*i.e.* the epidemic peak).

## Estimate the peak date

We'll estimate the peak date with the *incidence2* package.

```{r}
library(incidence2)
linelist$date_onset <- as.Date(linelist$onset, origin = "2020-01-01") 

incid <- incidence2::incidence(linelist, 
                               date_index = "date_onset", 
                               groups = "group")
peak <- estimate_peak(regroup(incid), progress = FALSE)$observed_peak
peak

```

## Mask non-direct transmissions

Group transmission assortativity is estimated from direct transmissions only. We can use the `filter_chain` function to mask non-direct transmissions (*i.e.* where `kappa`\>1).

```{r}
out_direct <- filter_chain(
  out,
  param = "kappa",
  thresh = 1,
  comparator = "==",
  target = "alpha" # keep alpha values where kappa == 1
)

trees <- get_trees(out_direct,
                   group = linelist$group,
                   onset = linelist$date_onset)

# keeps only rows where both onset dates are on or before peak
cut_trees <- lapply(trees, function(tree) {
  tree |> filter(from_onset <= peak, to_onset <= peak)
})

```

## Estimate group transmission assortativity

**To run the below you need to install *linktree* from [GitHub](https://github.com/CyGei/linktree/).**

For a given tree, we can estimate the groups' transmission assortativity coefficients with:

```{r, eval = requireNamespace("linktree", quietly = TRUE)}
library(linktree)
# staff-to-patient ratio
ratio <- c("hcw" = 1, "patient" = 3) 

max_post <- trees[[which.max(out$post)]]
delta <- linktree::get_delta(from = max_post$from_group, to = max_post$to_group, f = ratio)
plot(delta)
```

To account for uncertainty in who infected whom, we can compute the posterior distribution of assortativity coefficients. This is done by estimating `delta` for each tree in the posterior sample.

```{r, eval = requireNamespace("linktree", quietly = TRUE)}
deltas <- lapply(trees, function(x) {
  linktree::get_delta(from = x$from_group, to = x$to_group, f = ratio)
}) |>
  bind_rows(.id = "tree")

ggplot(deltas) +
  geom_density(aes(x = est, fill = group), bw = 0.03, color = "white") +
  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.2)) +
  scale_fill_manual(values = c("hcw" = "orange", "patient" = "purple")) +
  theme_classic() +
  theme(legend.position = c(0.1, 0.85))
```

Convert `delta` to `gamma` for interpretatibilty:

```{r, eval = requireNamespace("linktree", quietly = TRUE)}
gammas <- deltas |> 
  group_by(group) |>
  summarise(
    mean = mean(est),
    q025 = quantile(est, 0.025),
    q975 = quantile(est, 0.975)
  ) |>
  mutate(across(.cols = -group, ~ linktree::delta2gamma(.x)))

head(gammas)
```

Healthcare workers are nearly 5 times more likely to infect other HCWs whereas patients are 2.6 (`1/0.387`) times more likely to infect other HCWs.
